---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Welcome to the `roamR` package

<!-- badges: start -->
<!-- badges: end -->

`roamR` is an R package that offers a toolset and framework for running Individual Based Models (IBMs - or Agent-Based Models). It was created as part of the DisNBS project (Distribution of seabirds in the Non-Breeding Season) which was a project within the Off-shore Renewables Joint Industry Partnership programme ([ORJIP](http://www.orjip.org.uk/)). While the project has specific goals in relation to Off-shore Wind Farms (OWF) and their potential effects on seabirds, the problem is quite general. This package can help build animal movement simulations, with the general problem broken down into logical components and imposes a level of transparent formality, such that the computational components are on a common footing, leaving focus on parametrisation and inputs/data.

`roamR` is written in S4 and has been structured/implemented such that the simulations are well defined, but with a lot of flexibility in what the components can do. Animal IBMs are spatio-temporal and can be based on very disparate data sources - so `roamR` is particular in its treatment of space (projections etc), time and units generally (as found in the `units` package). The package was written with animals in mind (indeed one of the main classes is `Species`), but we'll use the agnostic term __agent__.

In brief - `roamR`:

- defines the environment the animal interacts with, and how this functionally affects its movement or behaviour
- defines how animals move, with consideration of the environment, behavoural state, or condition
- defines the behaviours animals may engage in and how these change
- defines the general characteristics of a type of agent (`class:Species`)
- creates many agents with stochastic properties in line with their species definition
- runs the agents through the simulated environment over time, for a defined temporal resolution
- measures the agents throughout the simulation
- gives tools to query the mass of agents that have been simulated

A principal limitation in IBMs is the level of information you have about how the agents behave. `roamR` is intended to work with very little information, up to very information rich. The 



## Installation

You can install the development version of roamR like so:

``` r
# FILL THIS IN! HOW CAN PEOPLE INSTALL YOUR DEV PACKAGE?
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(roamR)
## basic example code
```

