---
title: "IBM initiatization Walk-through"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IBM initiatization Walk-through}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(roamR)
```

This article is provides a step-by-step guide on how to initiate the
Individual-Based Model (IBM) underpinning the `{roamR}` package. The
initialization function `rmr_initiate()` requires the following objects:

- `<Species>`, containing the species-level properties of the agents

- `<Habitat>`, comprising the geographical and environmental features 
of the simulated area, from here on referred to as the Area of Calculation (AOC)

- a list of `<Structure>` objects, listing existent or future man-made physical 
structures that influence the simulated history of the agents

- `<ModelConfig>`, the main simulation settings





## Create the `<Species>` object

Arguments required 
