---
title: "Getting Started"
vignette: >
  %\VignetteIndexEntry{roamR}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

*Page Under Construction*


```{r}
library(roamR)
```

## Introduction

This article provides a fast-track hands-on demonstration of how to build an
Individual-Based Model (IBM) using roamR. We'll walk through a mock example
featuring an imaginary population in a hypothetical habitat to illustrate the
core workflow. In this example, we'll use the **Density-informed** movement
model.

The aim is to help you quickly familiarise yourself with the package's main
functions, object classes, and features, while providing a practical reference
for future implementations. For a more detailed description of roamR's
components, see the [roamR Overview](articles/roamR-guide.html) article.

## General roamR workflow

The standard workflow to construct and run an IMB in roamR can be summarised in
the following steps:

1.  **High-level model configuration** - Set the spatial and temporal boundaries
and resolutions, number of agents, and other simulation-level parameters.

2.  **Specify species features** - Define attributes such as initial body mass,
flight speed, activity states and their energetic trade-offs, and interactions
with environment.

3.  **Define the environment** - Include relevant spatio-temporal layers within
the simulated area such as species distributions, coastlines, man-made
structures, environmental attributes, etc.

4.  **Combine components into an IBM object** - Integrate the species and
environment definitions into a single simulation-ready object.

5.  **Run the simulation** - Create agents, parallelise computations, move
agents through the environment, and record their properties over time -
optionally including defined perturbation scenarios.

6.  **Query simulated animals** - Investigate results, e.g. by comparing
individual responses across baseline and impact scenarios.

In the following sections, we'll go over each of these steps and build our own
IBM from scratch. Buckle up!



## The *Turtwick* IBM: the baseline scenario

In this example, we focus on a hypothetical species, the
*Turtwick*, which inhabits grassland areas expected to be altered by the
introduction of man-made structures. Previous observations indicate that
*turtwicks* tend to avoid such structures, often showing displacement behaviour
in response to their presence.

We also know that the species' body mass and activity behaviours (here referred
to as **states**) are influenced by surface temperature and vegetation index.
*Turtwicks* are diurnal: individuals remain completely inactive at night,
resting in a nearby safe location once the sun sets.


### 1. Model Configuration

We start by defining the high-level configurations of the IBM for our *turtwick*
population using the `ModelConfig()` function:

```{r}
turtwick_ibm_cfg <- ModelConfig(
  movement_type = "di",
  n_agents = 100,
  ref_sys = sf::st_crs(32630), 
  aoc_bbx = c(0, 0, 10000, 10000), 
  delta_time = "1 day", 
  start_date = as.Date("2025-01-01"), 
  end_date = as.Date("2025-02-01")
)

turtwick_ibm_cfg
```

In plain terms, this configuration sets up a model where we:

- Simulate the movement and life history of **100** *turtwicks* (`n_agents`)
from 1 January to 1 February 2025, using daily time steps (`delta_time`,
`start_date`, `end_date`).

- Use the **UTM 30N** coordinate reference system (`ref_sys`).

- Restrict movement to a **10x10 km** area of calculation (AOC) with its origin at
**(0,0)**, using a spatial resolution of **250x250 m** cells (`delta_x` and
`delta_y`)

- Initialise agents at random locations within the AOC, allowing them to end up 
anywhere inside this area by the end of the simulation.




### 2. Species Definition

babababa bbababa maldodsdf sjaf kfodhf OHDFOH FOhf ofh ohfdhf odhfo godhgo goshg
[ojf j] pdsgj pgjpjgpdojg pjg 


### 2. Environment Specification

babababa bbababa maldodsdf sjaf kfodhf OHDFOH FOhf ofh ohfdhf odhfo godhgo goshg
[ojf j] pdsgj pgjpjgpdojg pjg 



## Disturbing *Turtwegs*: baseline Vs impact scenarios
