---
title: "guillemot-IoM-walkthrough"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tidyverse)
library(roamR)
```


# Overview

Example presented here will cover one species (Guillemot) focussing on the Isle of May (IoM) area. Only one agent will be run for example purposes.


* Simulation controls
* Environment and impact inputs
* Species-level inputs
* Agent-level inputs

```{r}


library(roamR)
library(stars)

```


# Initialisation

Overarching class for the IBM is `IBM` with slots for agents, drivers, species and configuration settings. For the purposes of this example, only some of the slots will be populated... 


```{r}
library(sf)
library(dplyr)
library(distributional)

# IBM settings -------------------------------------------------

isle_may <- st_sf(
  id = "Isle of May",
  prop = 1,
  geom = st_sfc(st_point(c(-2.5667, 56.1833))),
  crs = 4326
)

# UTM zone 30N
utm30 <- st_crs(32630)

# IBM Settings - modify parameter values to your specific case
guill_imb_config <- ModelConfig(
  n_agents = 1,
  ref_sys = utm30,
  aoc_bbx = isle_may |> st_transform(utm30) |> st_buffer(2.5e5) |>  st_bbox(), #
  delta_x = 1000,
  delta_y = 1000,
  time_step = "1 day",
  start_date = Sys.Date(),
  end_date = Sys.Date() + 180, # ~6 months
  start_sites = isle_may |> st_transform(utm30)
)


```


## Simulation settings

* Area of Calculation (AoC): a bounding box over which the simulation will run. Defines a repulsive boundary for agents. This ought to roughly exclude areas where the simulated population are not expected to be found.
* Number of agents
* Time-step resolution
* Simulation duration
* Nominal starting date/time 


We're going to use the following density map that governs Isle of May (IoM) guillemots for a specific month - noting these are bootstrap realisations from BioSS. From this we can infer an AoC and see the underlying coastal components within this. A single density map for month 3 (inferred from file delivery).


```{r}

density_map <- readRDS("data/GuillemotIsle of May3_iteration1.rds") %>% 
  st_as_stars() %>% 
  rename(pop_dens = X3) 

plot(density_map)

```

For reference and use later, an example competitors map:

```{r}

competitor_map <- readRDS("data/competition_layer_conmap_UK9004171_3.rds") %>% 
  st_as_stars() %>% 
  rename(compet_dens = layer.1) %>% 
  st_transform(st_crs(utm30))

plot(competitor_map)

```



This is a pretty big region - limit this to a smaller area for the purposes of this example. The bounding box is a requirement from the user.



```{r}

AoC <- guill_imb_config@aoc_bbx 

#%>% 
#aoc_bbxst_bbox(c(xmin = -5e5, ymin = -4e6,  xmax = 5e5, ymax = -3.2e6), crs = utm30) %>% 
  st_as_stars()

density_map <- st_crop(density_map, AoC)


competitor_map <- st_crop(competitor_map, AoC)

IFD_map <- (density_map + competitor_map) %>% 
  rename(combined_dens = pop_dens)

footprints <- st_read("data/Synthetic Polygons/") %>% 
  st_transform(st_crs(density_map)) %>% 
  st_crop(AoC)

coastline <- st_read("data/coastline/") %>% 
  st_transform(st_crs(density_map)) %>% 
  st_crop(AoC) 


ggplot(coastline) + 
  viridis::scale_fill_viridis() +
  geom_stars(data = IFD_map, aes(fill = combined_dens), alpha = 0.3) +
  geom_sf() + 
  geom_sf(data = footprints, fill = "lightblue") 
```


```{r}

sst_raw <- stars::read_ncdf("data/sst.mnmean.nc")


sst_drv <- Driver(
  id = "SST",
  type = "habitat",
  descr = "Sea Surface Temperature",
  stars_obj = stars::read_ncdf("data/sst.mnmean.nc") %>% 
    stars::st_as_stars() %>% 
    st_crop(AoC)
)


```


```{r}
library(sf)
library(dplyr)
library(distributional)

# IBM settings -------------------------------------------------

isle_may <- st_sf(
  id = "Isle of May",
  prop = 1,
  geom = st_sfc(st_point(c(-2.5667, 56.1833))),
  crs = 4326
)

# UTM zone 30N
utm30 <- st_crs(32630)
```


```{r}
# IBM Settings - modify parameter values to your specific case
guill_imb_config <- ModelConfig(
  n_agents = 1,
  ref_sys = utm30,
  aoc_bbx = isle_may |> st_transform(utm30) |> st_buffer(2.5e5) |>  st_bbox(), #
  delta_x = 1000,
  delta_y = 1000,
  time_step = "1 day",
  start_date = Sys.Date(),
  end_date = Sys.Date() + 180, # ~6 months
  start_sites = isle_may |> st_transform(utm30)
)
```




```{r}
# States specification -------------------------------------------------

states <- list(
  flight = State(
    id = "flight",
    energy_cost = VarDist(dist_normal(141, 66), units = "kJ/hour/gram"), # vals from excel sheet. Normal not adequate as it will generate negative values - needs reparameterization to logN
    time_budget = VarDist(dist_uniform(0, 0.15), "hours/day"), # roughly based on Lila's summaries
    speed = VarDist(dist_uniform(10, 20), "m/s") # made-up
  ),
  dive = State(
    id = "diving",
    energy_cost = VarDist(dist_normal(200, 10), units = "kJ/hour/gram"), # made-up
    time_budget = VarDist(dist_uniform(2.5, 3.5), "hours/day"), # roughly based on Lila's summaries
    speed = VarDist(dist_uniform(10, 20), "m/s") # made-up
  ),
  active = State(
    id = "active_on_water",
    energy_cost = VarDist(dist_normal(150, 10), units = "kJ/hour/gram"), # made-up
    time_budget = VarDist(dist_uniform(10, 11), "hours/day"), # roughly based on Lila's summaries
    speed = VarDist(dist_uniform(0, 2), "m/s") # made-up
  ),
  inactive = State(
    id = "inactive_on_water",
    energy_cost = VarDist(dist_normal(150, 10), units = "kJ/hour/gram"), # made-up
    time_budget = VarDist(dist_uniform(10, 11), "hours/day"), # roughly based on Lila's summaries
    speed = VarDist(dist_uniform(0, 1), "m/s") # made-up
  ),
  colony = State(
    id = "colony",
    energy_cost = VarDist(dist_normal(33.8, 11.4), units = "kJ/hour/gram"), # vals from excel sheet
    time_budget = VarDist(0, "hours/day"), # no time at colony
    speed = VarDist(0, "m/s") # made-up
  )
)



# Response to drivers  -------------------------------------------------

driver_resp <- list(
  coast = DriverResponse(
    driver_id = "land",
    movement = MoveInfluence(
      prob = VarDist(1),
      fn = function(x, slope = 1/2) exp(-slope * x),
      type = "repulsion"
    )),
  owf = DriverResponse(
    driver_id = "owf_foot",
    movement = MoveInfluence(
      prob = VarDist(dist_beta(2, 8)), # made-up
      fn = function(x, slope = 1/2) exp(-slope * x),
      type = "repulsion"
    ))#,
  # dens = DriverResponse(
  #   driver_id = "dens",
  #   movement = MoveInfluence(
  #     prob = VarDist(1),
  #     type = "attraction"
  #   )),
  # conspec = DriverResponse(
  #   driver_id = "conspec",
  #   movement = MoveInfluence(
  #     prob = VarDist(1),
  #     type = "repulsion"
  #   ))
)



# Specify <Species>  -------------------------------------------------

guill <- Species(
  id = "guill",
  common_name = "guillemot",
  scientific_name = "Uria Aalge",
  pop_size = 15000,
  body_mass_distr = VarDist(dist_uniform(600, 1450), "g"),
  mortality_thresh_distr = VarDist(800, "g"),
  states_profile = states,
  driver_responses = driver_resp
)


# Initiate Agent  -------------------------------------------------
simBird_1 <- Agent(species = guill, model_config = guill_imb_config)








# FOR REFERENCE: Initiate IBM -------------------------------------------------

guill_drivers <- list(
  drv_land = Driver(
    id = "land",
    type = "habitat",
    descr = "land",
    sf_obj = uk_land |> st_transform(utm30)
  ),
  drv_owfs = Driver(
    id = "owf_foot",
    type = "impact",
    descr = "OWF Footprints",
    sf_obj = owf_foots |> st_transform(utm30)
  ),
  drv_dens = Driver(
    id = "dens",
    type = "habitat",
    descr = "Guillemot spatial distribution",
    stars_obj = stars::st_as_stars(matrix(NA))  # plugin appropriate raster
  ),
  drv_conspec = Driver(
    id = "conspec",
    type = "habitat",
    descr = "conspecific distribution",
    stars_obj = stars::st_as_stars(matrix(NA)) # plugin appropriate raster
  )
)


guill_ibm <- rmr_initiate(
  species = guill,
  drivers = guill_drivers,
  model_config = guill_imb_config
)


```



```{r}

current_time <- now()
tz(current_time) <- "UTC"

my_config <- ModelConfig(n_agents = 1L,
                        ref_sys = sf::st_crs(32631), #UTM 31N
                        #ref_sys = sf::st_crs(4326), 
                        aoc_bbx = c(0, 0, 10, 10),
                        delta_x = 1000, #m 
                        delta_y = 1000,
                        time_step = "1 day",
                        start_date = Sys.Date() - 5,
                        end_date = Sys.Date(),
                        start_sites = NULL,
                        end_sites = NULL
)

```


```{r}

density_map <- st_transform(density_map, my_config@ref_sys)
AoC <- st_transform(AoC, my_config@ref_sys)

my_config@aoc_bbx <- AoC


aoc_grid <- sf::st_make_grid(
    my_config@aoc_bbx,
    cellsize = c(my_config@delta_x, my_config@delta_y),
    what = "centers"
  ) %>% 
  st_sf()

```


```{r}

aoc_driver <- generate_aoc_driver(grid = aoc_grid, bbox = AoC)

```




## Species definitions

* Behaviour profiles
* Speed and turning variance
* ...

state_expenditure <- c(141, 33.8, 113-(2.75*SST), 72.2-(2.75*SST), 3.71*(sum(1-exp(-dive_duration/1.23))))

1. Start by setting the behaviour profile

   Behaviour `mean(Hours)`
   <chr>             <dbl>
 1 Active          10.5
 2 Colony           0.0627
 3 Dive             3.11
 4 Flight           0.0559
 5 Inactive        10.3



```{r}

# 1. Start by setting the behaviour profile

SST <- 15
t_dive <- 5

behav <- list(
 flight = BehaviourSpec(
    behav = "flying",
    energy_cost = VarDist(dist_normal(141, 66), units = "kJ/hour/gram"),
    time_budget = VarDist(dist_uniform(0.056, 0.056), "hours/day"),
    speed = VarDist(dist_gamma(shape = 8, rate = 1/0.5), "m/s")
    ),
 dive = BehaviourSpec(
    behav = "diving",
    energy_cost = VarDist(dist_normal(3.71*(3.11*60/t_dive*(1-exp(-5/1.23))), 1.3), units = "kJ/min/gram"), # note units are per minute
    time_budget = VarDist(dist_uniform(3.11, 3.11), "hours/day"),
    speed = VarDist(dist_uniform(0, 1), "m/s")
  ),
 active = BehaviourSpec(
    behav = "swimming",
    energy_cost = VarDist(dist_normal(113-(2.75*SST), 22), units = "kJ/hour/gram"),
    time_budget = VarDist(dist_uniform(10.5, 10.5), "hours/day"),
    speed = VarDist(dist_uniform(0, 1), "m/s")
  ),
 inactive = BehaviourSpec(
    behav = "other",
    energy_cost = VarDist(dist_normal(72.2-(2.75*SST), 22), units = "kJ/hour/gram"),
    time_budget = VarDist(dist_uniform(10.3, 10.3),"hours/day"),
    speed = VarDist(dist_uniform(0, 0.1), "m/s")
  ),
 colony = BehaviourSpec(
    behav = "nest_attending",
    energy_cost = VarDist(dist_normal(33.8, 11.4), units = "kJ/hour/gram"),
    time_budget = VarDist(dist_uniform(0, 0), "hours/day"),
    speed = VarDist(dist_uniform(0, 0), "m/s")
  )
)

```



## Environment, drivers and impacts

* 

```{r}
# 2. the list of impact responses
imp_resp <- list(
  owf_footprint = ImpactResponse(
    impact_id = "owf_foot",
    displace_prob = VarDist(dist_beta(2, 8)),
    displace_fn = function(x, y){ x/y},
    disturb_prob = VarDist(dist_beta(1, 5)),
    disturb_behav = "flying",
    # extra hours/day of flight due to impact
    disturb_extent = VarDist(dist_lognormal(2, 1), "hr/day")
  )
)

# 3. `stars` array eith 10 samples of monthly density distribution
dens <- data.frame(
           expand.grid(x= 1:5, y = 1:5, month = 3:5, iter = as.integer(1:10)),
           counts = rlnorm(5*5*3*10)) |>
        st_as_stars(dims = c("x", "y", "month", "iter"))

# 4. specify <Species>
guill <- Species(
     id = "guill",
     role = "agent",
     common_name = "guillemot",
     scientific_name = "Uria Aalge",
     body_mass_distr = VarDist(dist_uniform(850, 1130), "g"),
     mortality_thresh_distr = VarDist(dist_normal(500, 20), "g"),
     spatial_distr = dens,
     behaviour_profile = behav,
     impact_responses = imp_resp
   )


```



## Data

* Transition matrices - scaling for time-steps

```{r}
#| eval: FALSE

lila_dat <- read_csv("data/ActivityBudget.csv")

lila_mat <- lila_dat %>% 
  group_by(Colony, TDR_ID, Date, Behaviour) %>%
  summarise(sum(Hours))

```



## Parameters

# Running a simulation


```{r}

# agent

```


# Querying the outputs

